---
title: "Construction of 5'UTR Ancestral Sequences"
author: "Hailey Bruzzone"
date: "Septmber, 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/5-UTR")
library(phylotools)
```

## Reconstruct Various Ancestral Sequences using RAxML 

## Terminal Directory 

`cd /Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/5-UTR`

## IRES 1 ##

Trimmed:
1. MFP tree: SYM+G4

Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/SYM+G4/5UTR_MUSCLE_ires1_NT_hmm.fasta --tree ./data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/SYM+G4/muscle_hmm_sym_5utr_nt_ires1_rooted.treefile --msa-format FASTA --data-type DNA --model SYM+G4`

Output file is a table, with ancestral sequence labeled as NA, so we convert it to a FASTA file and insert "1" into the NA entry to indicate its place in ancestry

```{r trimmed sym ires1}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/haileylouw/Desktop/ViralAncestral/results/raxml-ng")
library(phylotools)
fiveUTR_muscle_ires1_nt_hmm_sym_ancestral <- read.table("./data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/SYM+G4/5UTR_MUSCLE_ires1_NT_hmm.fasta.raxml.ancestralStates")
colnames(fiveUTR_muscle_ires1_nt_hmm_sym_ancestral) <- c("seq.name", "seq.text")
fiveUTR_muscle_ires1_nt_hmm_sym_ancestral[20,1] <- 1

## Write to directory
dat2fasta(fiveUTR_muscle_ires1_nt_hmm_sym_ancestral, "./data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/SYM+G4/fiveUTR_muscle_ires1_nt_hmm_sym_ancestral.fasta")
```

2. GTR+I+G4 

Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires1_NT_hmm.fasta --tree ./data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/GTR+I+G4/muscle_hmm_gtr_5utr_nt_ires1_rooted.treefile --msa-format FASTA --data-type DNA --model GTR+I+G4`

Output file is a table, with ancestral sequence labeled as NA, so we convert it to a FASTA file and insert "1" into the NA entry to indicate its place in ancestry

```{r trimmed gtr ires1}
fiveUTR_muscle_ires1_nt_hmm_gtr_ancestral <- read.table("~/Desktop/ViralAncestral/results/raxml-ng/data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires1_NT_hmm.fasta.raxml.ancestralStates")
colnames(fiveUTR_muscle_ires1_nt_hmm_gtr_ancestral) <- c("seq.name", "seq.text")
fiveUTR_muscle_ires1_nt_hmm_gtr_ancestral[20,1] <- 1

## Write to directory
dat2fasta(fiveUTR_muscle_ires1_nt_hmm_gtr_ancestral, "~/Desktop/ViralAncestral/results/raxml-ng/data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/GTR+I+G4/fiveUTR_muscle_ires1_nt_hmm_gtr_ancestral.fasta")
```

Untrimmed: 
1. MFP tree: SYM+G4

Command: `./SYM+G4/raxml-ng/raxml-ng --ancestral --msa ./SYM+G4/IRES1_with_outgroup_aligned.fasta --tree ./SYM+G4/muscle_sym_5utr_nt_ires1_rooted.treefile --msa-format FASTA --data-type DNA --model SYM+G4`

Output file is a table, with ancestral sequence labeled as NA, so we convert it to a FASTA file and insert "1" into the NA entry to indicate its place in ancestry

```{r sym ires1}
fiveUTR_muscle_ires1_nt_sym_ancestral <- read.table("~/Desktop/ViralAncestral/results/raxml-ng/data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/SYM+G4/5UTR_MUSCLE_ires1_NT.fasta.raxml.ancestralStates")
colnames(fiveUTR_muscle_ires1_nt_sym_ancestral) <- c("seq.name", "seq.text")
fiveUTR_muscle_ires1_nt_sym_ancestral[20,1] <- 1

## Write to directory
dat2fasta(fiveUTR_muscle_ires1_nt_sym_ancestral, "~/Desktop/ViralAncestral/results/raxml-ng/data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/SYM+G4/fiveUTR_muscle_ires1_nt_sym_ancestral.fasta")
```
```{r untrimmed sym tree, echo=FALSE, message=FALSE, warning=FALSE}
library(phytools)
fiveUTR_IRES1_SYM_ancestral_state <- read.table("~/Desktop/ViralAncestral/results/5UTR_Trees/MFP/SYM+G4 IRES 1 Ancestral with Mitovirus Outgroup/IRES1_with_outgroup_aligned.fasta.raxml.ancestralStates")
fiveUTR_IRES1_SYM_ancestral_tree <- read.tree("~/Desktop/ViralAncestral/results/5UTR_Trees/MFP/SYM+G4 IRES 1 Ancestral with Mitovirus Outgroup/IRES1_with_outgroup_aligned.fasta.raxml.ancestralTree")
plotTree(fiveUTR_IRES1_SYM_ancestral_tree, ftype = "i", fize = 0.5, lwd = 1)
labelnodes(1:fiveUTR_IRES1_SYM_ancestral_tree$Nnode+Ntip(fiveUTR_IRES1_SYM_ancestral_tree),
           1:fiveUTR_IRES1_SYM_ancestral_tree$Nnode+Ntip(fiveUTR_IRES1_SYM_ancestral_tree),
           interactive = FALSE, cex=0.3, shape = "ellipse")
```

2. GTR + I + G4 

Command: `./GTR+I+G4/raxml-ng/raxml-ng --ancestral --msa ./GTR+I+G4/IRES1_with_outgroup_aligned.fasta --tree ./GTR+I+G4/muscle_untrimmed_gtr_5utr_nt_ires1_mito_rooted.treefile --msa-format FASTA --data-type DNA --model GTR+I+G4`

Output file is a table, with ancestral sequence labeled as NA, so we convert it to a FASTA file and insert "1" into the NA entry to indicate its place in ancestry

```{r gtr ires1}
fiveUTR_muscle_ires1_nt_gtr_ancestral <- read.table("./data/5UTR_IRES1_MUSCLE_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires1_NT.fasta.raxml.ancestralStates")
colnames(fiveUTR_muscle_ires1_nt_gtr_ancestral) <- c("seq.name", "seq.text")
fiveUTR_muscle_ires1_nt_gtr_ancestral[20,1] <- 1

## Write to directory
dat2fasta(fiveUTR_muscle_ires1_nt_gtr_ancestral, "./data/5UTR_IRES1_MUSCLE_Ancestral/GTR+I+G4/fiveUTR_muscle_ires1_nt_gtr_ancestral.fasta")
```

Untrimmed, with mitovirus outgroup:

First need to number sequences / differentiate names for RAxML

```{r add nos, echo=FALSE, message=FALSE, warning=FALSE}
library(phytools)
fiveUTR_IRES1_GTR_ancestral_state <- read.table("~/Desktop/ViralAncestral/results/5UTR_Trees/GTR+F+I+G4/GTR+I+G4 IRES 1 Ancestral with Mitovirus Outgroup/IRES1_with_outgroup_aligned.fasta.raxml.ancestralStates")
fiveUTR_IRES1_GTR_ancestral_tree <- read.tree("~/Desktop/ViralAncestral/results/5UTR_Trees/GTR+F+I+G4/GTR+I+G4 IRES 1 Ancestral with Mitovirus Outgroup/IRES1_with_outgroup_aligned.fasta.raxml.ancestralTree")
plotTree(fiveUTR_IRES1_GTR_ancestral_tree, ftype = "i", fize = 0.5, lwd = 1)
labelnodes(1:fiveUTR_IRES1_GTR_ancestral_tree$Nnode+Ntip(fiveUTR_IRES1_GTR_ancestral_tree),
           1:fiveUTR_IRES1_GTR_ancestral_tree$Nnode+Ntip(fiveUTR_IRES1_GTR_ancestral_tree),
           interactive = FALSE, cex=0.3, shape = "ellipse")
```

1. MFP tree: SYM+G4
Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES1_MUSCLE_Ancestral_Mito/IRES1_with_outgroup_aligned.fasta --tree ./data/5UTR_IRES1_MUSCLE_Ancestral_Mito/SYM+G4/IRES1_with_outgroup_aligned.fasta.treefile --msa-format FASTA --data-type DNA --model SYM+G4`

2. GTR + I + G4 
Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES1_MUSCLE_Ancestral_Mito/IRES1_with_outgroup_aligned.fasta --tree ./data/5UTR_IRES1_MUSCLE_Ancestral_Mito/GTR+I+G4/IRES1_with_outgroup_aligned.fasta.treefile --msa-format FASTA --data-type DNA --model GTR+I+G4`


### IRES 2

Trimmed:
1. MFP tree: TIM3e+I+G4

Note that this evolutionary model is not supported by RAxML at the moment, but its command would follow a similar pattern to the other sequences, with the TIM3e+I+G4 evolutionary model specified. 

2. GTR + I + G4 

Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES2_MUSCLE_Trimmed_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires2_NT_hmm.fasta --tree ./data/5UTR_IRES2_MUSCLE_Trimmed_Ancestral/GTR+I+G4/muscle_hmm_gtr_5utr_nt_ires2_rooted.treefile --msa-format FASTA --data-type DNA --model GTR+I+G4`

```{r trimmed gtr ires1}
fiveUTR_muscle_ires2_nt_hmm_gtr_ancestral <- read.table("./data/5UTR_IRES2_MUSCLE_Trimmed_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires2_NT_hmm.fasta.raxml.ancestralStates")
colnames(fiveUTR_muscle_ires2_nt_hmm_gtr_ancestral) <- c("seq.name", "seq.text")

## Write to directory
dat2fasta(fiveUTR_muscle_ires2_nt_hmm_gtr_ancestral, "./data/5UTR_IRES2_MUSCLE_Trimmed_Ancestral/GTR+I+G4/fiveUTR_muscle_ires2_nt_hmm_gtr_ancestral.fasta")
```

Untrimmed: 
1. MFP tree: TPM3u+F+R4

Note that this evolutionary model is not supported by RAxML at the moment, but its command would follow a similar pattern to the other sequences, with the TPM3u+F+R4 evolutionary model specified. 

2. GTR+I+G4 

Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES2_MUSCLE_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires2_NT.fasta --tree ./data/5UTR_IRES2_MUSCLE_Ancestral/GTR+I+G4/muscle_gtr_5utr_nt_ires2_rooted.treefile --msa-format FASTA --data-type DNA --model GTR+I+G4`

```{r gtr ires1}
fiveUTR_muscle_ires2_nt_gtr_ancestral <- read.table("./data/5UTR_IRES2_MUSCLE_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires2_NT.fasta.raxml.ancestralStates")
colnames(fiveUTR_muscle_ires2_nt_gtr_ancestral) <- c("seq.name", "seq.text")

## Write to directory
dat2fasta(fiveUTR_muscle_ires2_nt_gtr_ancestral, "./data/5UTR_IRES2_MUSCLE_Ancestral/GTR+I+G4/fiveUTR_muscle_ires2_nt_gtr_ancestral.fasta")
```

### IRES 3 

Trimmed:
1. MFP tree: HKY+F+G4

Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES3_MUSCLE_Trimmed_Ancestral/HKY+F+G4/5UTR_MUSCLE_ires3_NT_hmm.fasta --tree ./data/5UTR_IRES3_MUSCLE_Trimmed_Ancestral/HKY+F+G4/muscle_hmm_hky_5utr_nt_ires3_rooted.treefile --msa-format FASTA --data-type DNA --model HKY+F+G4`

```{r trimmed sym ires1}
fiveUTR_muscle_ires3_nt_hmm_hky_ancestral <- read.table("./data/5UTR_IRES3_MUSCLE_Trimmed_Ancestral/HKY+F+G4/5UTR_MUSCLE_ires3_NT_hmm.fasta.raxml.ancestralStates")
colnames(fiveUTR_muscle_ires3_nt_hmm_hky_ancestral) <- c("seq.name", "seq.text")

## Write to directory
dat2fasta(fiveUTR_muscle_ires3_nt_hmm_hky_ancestral, "./data/5UTR_IRES3_MUSCLE_Trimmed_Ancestral/HKY+F+G4/fiveUTR_muscle_ires3_nt_hmm_hky_ancestral.fasta")
```

2. GTR+I+G4 

Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES3_MUSCLE_Trimmed_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires3_NT_hmm.fasta --tree ./data/5UTR_IRES3_MUSCLE_Trimmed_Ancestral/GTR+I+G4/muscle_hmm_gtr_5utr_nt_ires3_rooted.treefile --msa-format FASTA --data-type DNA --model GTR+I+G4`

```{r trimmed gtr ires1}
fiveUTR_muscle_ires3_nt_hmm_gtr_ancestral <-read.table("./data/5UTR_IRES3_MUSCLE_Trimmed_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires3_NT_hmm.fasta.raxml.ancestralStates")
colnames(fiveUTR_muscle_ires3_nt_hmm_gtr_ancestral) <- c("seq.name", "seq.text")

## Write to directory
dat2fasta(fiveUTR_muscle_ires3_nt_hmm_gtr_ancestral, "./data/5UTR_IRES3_MUSCLE_Trimmed_Ancestral/GTR+I+G4/fiveUTR_muscle_ires3_nt_hmm_gtr_ancestral.fasta")
```

Untrimmed: 
1. MFP tree: TN+F+G4

Note that this evolutionary model is not supported by RAxML at the moment, but its command would follow a similar pattern to the other sequences, with the TN+F+G4 evolutionary model specified. 

2. GTR+I+G4 

Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES3_MUSCLE_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires3_NT.fasta --tree ./data/5UTR_IRES3_MUSCLE_Ancestral/GTR+I+G4/muscle_gtr_5utr_nt_ires3_rooted.treefile --msa-format FASTA --data-type DNA --model GTR+I+G4`

```{r gtr ires1}
fiveUTR_muscle_ires3_nt_gtr_ancestral <- read.table("./data/5UTR_IRES3_MUSCLE_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires3_NT.fasta.raxml.ancestralStates")
colnames(fiveUTR_muscle_ires3_nt_gtr_ancestral) <- c("seq.name", "seq.text")

## Write to directory
dat2fasta(fiveUTR_muscle_ires3_nt_gtr_ancestral, "./data/5UTR_IRES3_MUSCLE_Ancestral/GTR+I+G4/fiveUTR_muscle_ires3_nt_gtr_ancestral.fasta")
```
