---
title: "Construction of 5'UTR Ancestral Sequences"
author: "Hailey Bruzzone"
date: "October 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/5-UTR")
library(phylotools)
```

## Reconstruct Various Ancestral Sequences using RAxML 

## Terminal Directory 

`cd /Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/5-UTR`

## IRES 1 5'UTR ##

Untrimmed NT: 

1. MFP tree: SYM+G4

Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/SYM+G4/5UTR_MUSCLE_ires1_NT_hmm.fasta --tree ./data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/SYM+G4/muscle_hmm_sym_5utr_nt_ires1_rooted.treefile --msa-format FASTA --data-type DNA --model SYM+G4`

Output file is a table, with ancestral sequence labeled as NA, so we convert it to a FASTA file and insert "1" into the NA entry to indicate its place in ancestry

```{r untrimmed sym ires1}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/5-UTR/SYM+G4")
library(phylotools)

# read in file 
fiveUTR_sym_ancestral <- read.table("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/5-UTR/SYM+G4/IRES1_with_outgroup_aligned.fasta.raxml.ancestralStates")
colnames(fiveUTR_sym_ancestral) <- c("seq.name", "seq.text")

## Write to directory
dat2fasta(fiveUTR_sym_ancestral, "/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/5-UTR/SYM+G4/fiveUTR_sym_ancestral.fasta")
```

2. GTR+I+G4 

Command: `./raxml-ng --ancestral --msa ./data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/GTR+I+G4/5UTR_MUSCLE_ires1_NT_hmm.fasta --tree ./data/5UTR_IRES1_MUSCLE_Trimmed_Ancestral/GTR+I+G4/muscle_hmm_gtr_5utr_nt_ires1_rooted.treefile --msa-format FASTA --data-type DNA --model GTR+I+G4`

```{r trimmed gtr ires1}
# read in file 
fiveUTR_gtr_ancestral <- read.table("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/5-UTR/GTR+I+G4/IRES1_with_outgroup_aligned.fasta.raxml.ancestralStates")
colnames(fiveUTR_gtr_ancestral) <- c("seq.name", "seq.text")

## Write to directory
dat2fasta(fiveUTR_gtr_ancestral, "/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/5-UTR/GTR+I+G4/fiveUTR_gtr_ancestral.fasta")
```

## IRES 1 Replicase Gene ##

Untrimmed NT: 

1. SYM+G4

Directory: `cd /Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene`

Command: `./SYM+G4/raxml-ng/raxml-ng --ancestral --msa ./SYM+G4/ires1_REP_NT_with_outgroup_aligned.fasta --tree ./SYM+G4/muscle_sym_rep_nt_ires1_rooted.treefile --msa-format FASTA --data-type DNA --model SYM+G4`

```{r untrimmed sym rep ires1}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene/SYM+G4")
library(phylotools)

# read in table 
rep_sym_ancestral <- read.table("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene/SYM+G4/ires1_REP_NT_with_outgroup_aligned.fasta.raxml.ancestralStates")
colnames(rep_sym_ancestral) <- c("seq.name", "seq.text")

## Write to directory
dat2fasta(rep_sym_ancestral, "/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene/SYM+G4/rep_sym_ancestral.fasta")
```

2. GTR+I+F+G4

Directory: `cd /Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene`

Command: `./GTR+I+G4/raxml-ng/raxml-ng --ancestral --msa ./GTR+I+G4/ires1_REP_NT_with_outgroup_aligned.fasta --tree ./GTR+I+G4/muscle_gtr_rep_nt_ires1_rooted.treefile --msa-format FASTA --data-type DNA --model GTR+I+G4`

```{r untrimmed gtr rep ires1}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene/GTR+F+I+G4")
library(phylotools)

# read in table 
rep_gtr_ancestral <- read.table("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene/GTR+F+I+G4/ires1_REP_NT_with_outgroup_aligned.fasta.raxml.ancestralStates")
colnames(rep_gtr_ancestral) <- c("seq.name", "seq.text")

## Write to directory
dat2fasta(rep_gtr_ancestral, "/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene/GTR+F+I+G4/rep_gtr_ancestral.fasta")
```

Untrimmed AA: 

3. LG (AA)

Directory: `cd /Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene`

Command: `./LG+F+I+G4/raxml-ng/raxml-ng --ancestral --msa ./LG+F+I+G4/ires1_REP_AA_with_outgroup_aligned.fasta --tree ./LG+F+I+G4/muscle_lg_rep_aa_ires1_rooted.treefile --msa-format FASTA --data-type AA --model LG+F+I+G4`

```{r untrimmed rep lg ires1}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene/LG+F+I+G4")
library(phylotools)

# read in table 
rep_lg_ancestral <- read.table("/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene/LG+F+I+G4/ires1_REP_AA_with_outgroup_aligned.fasta.raxml.ancestralStates")
colnames(rep_lg_ancestral) <- c("seq.name", "seq.text")

## Write to directory
dat2fasta(rep_lg_ancestral, "/Users/haileylouw/Desktop/ViralAncestral/Results_Fall_2024/Replicase_Gene/LG+F+I+G4/rep_lg_ancestral.fasta")
```


4. Q.yest (AA) --> can't do in RAxML right now 

Command: `./Q.yeast+R3/raxml-ng/raxml-ng --ancestral --msa ./Q.yeast+R3/ires1_REP_AA_with_outgroup_aligned.fasta --tree ./Q.yeast+R3/muscle_q_rep_aa_ires1_rooted.treefile --msa-format FASTA --data-type AA --model Q.yeast+R3`

